---
title: "Clinical Visit Insights"
author: "David Willsher"
output:
  html_document:
    css: style.css
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
---

> **Executive Summary**  
> This report demonstrates a scalable R-based automation pipeline for extracting, cleaning, validating, and summarising clinical visit data from a Snowflake warehouse. It highlights how structured patient data, including embedded JSON, can be transformed into usable insights. This workflow enhances operational efficiency, supports healthcare research, and ensures high-quality data governance.

# 1. Project Overview

The pipeline pulls data from a Snowflake-hosted table (`clinical_visits_2`), parses nested JSON `vitals`, applies QA checks, and produces department-level summaries and visual insights. All credentials are securely managed via the `keyring` package, and outputs are saved for downstream use or reporting.

---

# 2. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(jsonlite)
library(purrr)
library(readr)
library(ggplot2)
df <- read_csv("output/cleaned_clinical_visits.csv")
qa_issues <- read_csv("output/qa_issues.csv")
```

The data was extracted using an ODBC connection and saved as CSVs for reproducibility.

---

# 3. Data Structure & Parsing

Each row in the dataset represents a clinical visit. The `vitals` field originally contained a JSON string with multiple readings. This was parsed into structured `bp` (blood pressure) and `temp` (temperature) columns.

## Sample Rows

```{r}
head(df, 5)
```

---

# 4. QA Checks & Validation

The following checks were applied:

- Missing values in `age`, `gender`, `bp`, or `temp`
- Implausible values:  
  - `age < 0` or `age > 120`  
  - `temp < 35°C` or `temp > 42°C`

## QA Issues by Reason

```{r}
qa_issues %>%
  group_by(reason = case_when(
    is.na(age) ~ "Missing age",
    is.na(gender) ~ "Missing gender",
    is.na(bp) ~ "Missing BP",
    is.na(temp) ~ "Missing temp",
    TRUE ~ "Other"
  )) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

---

# 5. Department-Level Summary

This summary shows patient distribution across departments and admission rates.

```{r}
df %>%
  group_by(department, admitted) %>%
  summarise(
    total_patients = n(),
    avg_age = round(mean(age, na.rm = TRUE), 1),
    avg_temp = round(mean(temp, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(total_patients))
```

---

# 6. Visual Insights

## Patient Count by Department

```{r, fig.width=6, fig.height=4}
ggplot(df, aes(x = department)) +
  geom_bar(fill = "#4e79a7") +
  theme_minimal() +
  coord_flip() +
  labs(title = "Patient Count by Department", x = "Department", y = "Count")
```

## Temperature Distribution

```{r, fig.width=6, fig.height=4}
ggplot(df, aes(x = temp)) +
  geom_histogram(binwidth = 0.2, fill = "#f28e2b", color = "white") +
  theme_minimal() +
  labs(title = "Temperature Distribution", x = "Temperature (°C)", y = "Frequency")
```

---

# 7. Observations and Business Relevance

## Operational Efficiency
- Automatically processes raw clinical visit data with minimal manual effort
- Streamlined JSON parsing and QA allow for scalable report generation

## Clinical & Research Value
- Enables rapid triage or analysis of abnormal patient presentations
- Clean vitals can feed into downstream models (e.g., risk scoring, alerts)

## Data Governance
- Built-in checks flag bad data early, improving reliability of any decisions
- Uses secure credential storage (via `keyring`) for GDPR compliance and security

---

# 8. Reproducibility Notes

- This report was generated using a reproducible R pipeline stored in GitHub.
- Credentials are **not** stored in the script and must be provided via `keyring`.
- Outputs include:
  - `cleaned_clinical_visits.csv`
  - `qa_issues.csv`
  - `department_summary.csv`

---

> *Developed by David Willsher — showcasing scalable automation for clinical data insight generation.*

